---
output: github_document
title: "Crowd-Optimized R Functions"
author: "Dane Van Domelen <br> vandomed@gmail.com"
date: "`r Sys.Date()`"
bibliography: vignettes/bibliography.bib
nocite: | 
  @rcpp1, 
  @rcpp2,
  @rcpp3,
  @rfast,
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  message = FALSE
)

# Load packages
library("knitr")
library("crowdopt")
library("microbenchmark")

# Other options
options(digits = 4)
set.seed(123)
```


## Concept

The concept for this package is crowdsourced optimization for speed. Long term, the goal is to optimize enough base R functions that a typical R user can drastically speed up their scripts by simply loading **crowdopt** and replacing functions like *mean* and *sd* with *mean2* and *sd2*.

Anyone can contribute on [GitHub](https://github.com/vandomed/crowdopt). Once you contribute a function, you become a co-author for the package, and remain a co-author even if your function eventually gets replaced by a faster one developed by somebody else.

To contribute a function, all you need to do is add a .R file with the code and documentation for your function, using *roxygen2* syntax. You can use `mean2.R` as a template. Be sure to include an *Examples* section in which you demonstrate that your function is faster than the corresponding base R function.

In general, contributed functions should be substantially faster than the corresponding base R function, ideally for both small and large objects (e.g. vectors of length 5 and 500,000).


## List of functions

| Function         | Faster version of: | C++?
|:-----------------|:-------------------|---------------------------
| *mean2*          | *mean*             | Yes (for integer vectors)
| *weighted.mean2* | *weighted.mean*    | Yes
| *sd2*            | *sd*               | Yes
| *var2*           | *var*              | Yes
| *cov2*           | *cov*              | Yes
| *weighted.mean2* | weighted.mean      | Yes


## Proof of outperformance

### mean2

This is a faster version of the base R function *mean*, defined simply as `sum(x) / length(x)`. Speed advantage diminishes with `length(x)`.

```{r, fig.width = 6, fig.height = 5}
lengths <- c(3, 10, 100, 1000, 10000, 50000, 100000)
multiples1 <- multiples2 <- c()
for (ii in 1: 7) {
  x1 <- rpois(lengths[ii], lambda = 3)
  x2 <- rnorm(lengths[ii])
  medians <- summary(microbenchmark(mean(x1), mean2(x1), mean(x2), mean2(x2), 
                                    times = 250))$median
  multiples1[ii] <- medians[1] / medians[2]
  multiples2[ii] <- medians[3] / medians[4]
}
plot(1: 7, multiples1, type = "b", col = "blue", main = "mean2 vs. mean", 
     ylab = "Speed multiple", xlab = "Vector length", xaxt = "n", 
     ylim = c(0, max(c(multiples1, multiples2)) * 1.05))
points(1: 7, multiples2, type = "b", col = "red")
axis(side = 1, at = 1: 7, labels = lengths)
abline(h = 1)
legend("topright", legend = c("Integer x", "Numeric x"), lty = 1, 
       col = c("blue", "red"))
```

### weighted.mean2

This is a faster version of the base R function *weighted.mean*. For optimal speed, specify `x.integer = TRUE` and `w.integer = TRUE` if `x` and `w` are integer vectors.

```{r, fig.width = 6, fig.height = 5}
multiples1 <- multiples2 <- c()
for (ii in 1: 7) {
  x1 <- rpois(lengths[ii], lambda = 3)
  w1 <- rpois(lengths[ii], lambda = 1)
  x2 <- rnorm(lengths[ii])
  w2 <- rexp(lengths[ii])
  medians <- summary(microbenchmark(weighted.mean(x1, w2), 
                                    weighted.mean2(x1, w1, TRUE, TRUE),
                                    weighted.mean(x2, w2), 
                                    weighted.mean2(x2, w2),
                                    times = 250))$median
  multiples1[ii] <- medians[1] / medians[2]
  multiples2[ii] <- medians[3] / medians[4]
}
plot(1: 7, multiples1, type = "b", col = "blue", 
     main = "weighted.mean2 vs. weighted.mean", 
     ylab = "Speed multiple", xlab = "Vector length", xaxt = "n", 
     ylim = c(0, max(c(multiples1, multiples2)) * 1.05))
points(1: 7, multiples2, type = "b", col = "red")
axis(side = 1, at = 1: 7, labels = lengths)
abline(h = 1)
legend("topleft", legend = c("Integer x, w", "Numeric x, w"), lty = 1, 
       col = c("blue", "red"))
```

### sd2

This is a faster version of base R function *sd*. For optimal speed, use `integer = TRUE` if `x` is an integer vector and `integer = FALSE` otherwise. Note that for long vectors the function *Var* in **Rfast** is typically much faster than *sd2*.

```{r, fig.width = 6, fig.height = 5}
for (ii in 1: 7) {
  x1 <- rpois(lengths[ii], lambda = 3)
  x2 <- rnorm(lengths[ii])
  medians <- summary(microbenchmark(sd(x1), sd2(x1, TRUE), sd(x2), sd2(x2), 
                                    times = 250))$median
  multiples1[ii] <- medians[1] / medians[2]
  multiples2[ii] <- medians[3] / medians[4]
}
plot(1: 7, multiples1, type = "b", col = "blue", main = "sd2 vs. sd", 
     ylab = "Speed multiple", xlab = "Vector length", xaxt = "n", 
     ylim = c(0, max(c(multiples1, multiples2)) * 1.05))
points(1: 7, multiples2, type = "b", col = "red")
axis(side = 1, at = 1: 7, labels = lengths)
abline(h = 1)
legend("topright", legend = c("Integer x", "Numeric x"), lty = 1, 
       col = c("blue", "red"))
```

### var2

This is a faster version of base R function *var*. For optimal speed, use `integer = TRUE` if `x` is an integer vector and `integer = FALSE` otherwise. Note that for long vectors the function *Var* in **Rfast** is typically much faster than *sd2*.

```{r, fig.width = 6, fig.height = 5}
for (ii in 1: 7) {
  x1 <- rpois(lengths[ii], lambda = 3)
  x2 <- rnorm(lengths[ii])
  medians <- summary(microbenchmark(var(x1), var2(x1, TRUE), var(x2), var2(x2), 
                                    times = 250))$median
  multiples1[ii] <- medians[1] / medians[2]
  multiples2[ii] <- medians[3] / medians[4]
}
plot(1: 7, multiples1, type = "b", col = "blue", main = "var2 vs. var", 
     ylab = "Speed multiple", xlab = "Vector length", xaxt = "n", 
     ylim = c(0, max(c(multiples1, multiples2)) * 1.05))
points(1: 7, multiples2, type = "b", col = "red")
axis(side = 1, at = 1: 7, labels = lengths)
abline(h = 1)
legend("topright", legend = c("Integer x", "Numeric x"), lty = 1, 
       col = c("blue", "red"))

```


### cov2

This is a faster version of base R function *cov*. For optimal speed, use `integer = TRUE` if `x` and `y` are integer vectors and `integer = FALSE` otherwise.


```{r, fig.width = 6, fig.height = 5}
for (ii in 1: 7) {
  x1 <- rpois(lengths[ii], lambda = 3)
  y1 <- rpois(lengths[ii], lambda = 3)
  x2 <- rnorm(lengths[ii])
  y2 <- rnorm(lengths[ii])
  medians <- 
    summary(microbenchmark(cov(x1, y1), cov2(x1, y1, TRUE), 
                           cov(x2, y2), cov2(x2, y2), times = 250))$median
  multiples1[ii] <- medians[1] / medians[2]
  multiples2[ii] <- medians[3] / medians[4]
}
plot(1: 7, multiples1, type = "b", col = "blue", main = "cov2 vs. cov", 
     ylab = "Speed multiple", xlab = "Vector length", xaxt = "n", 
     ylim = c(0, max(c(multiples1, multiples2)) * 1.05))
points(1: 7, multiples2, type = "b", col = "red")
axis(side = 1, at = 1: 7, labels = lengths)
abline(h = 1)
legend("topright", legend = c("Integer x", "Numeric x"), lty = 1, 
       col = c("blue", "red"))

```


## References
